Quantum Algorithms for Portflolio Optimisation and Option pricing

### 2024-06-11 project/paper-v2.tex
submit paper

\begin{abstract}
Quantum computing has the promise to address computational problems that are considered intractable with our classical computational models. Nevertheless, due to the current limitations of the available quantum hardware, we are still years away from achieving a universal fault-tolerant quantum computer. In the interim, the development of quantum algorithms for specific problems and their implementation on near-term quantum devices is a promising approach. In this work, we review the most promising quantum algorithms for financial applications, focusing on Portfolio Optimisation and Option Pricing. We describe the ideas behind important quantum algorithms such as Variational Quantum Algorithms (VQAs) and Quantum Amplitude Estimation (QAE) and their theoretical performance. We provide an overview of the hardware available for their implementation and discuss the results 
obtained by some study cases, comparing them with their classical counterparts. Finally, we discuss the future prospects of these algorithms in the financial industry.
  \\
  \textbf{Keywords:} Quantum algorithms, Quantum approximate optimization algorithm, Quantum amplitude estimation, Portfolio optimisation, Quantum computing.
\end{abstract}

@@

In Sec. \ref{sec:hardware} we provide an overview of the hardware implementations of these algorithms. Finally, Sec. \ref{sec:discussion} concludes this work and discusses the future prospects of quantum algorithms in the financial industry.

@@

A set of investments $x_i$ (given as a fraction of the total budget or number of asset units) allocated for each $i$th considered asset, 
is then defined. Therefore, an optimal method for asset allocation aims to maximise the portfolio return 
$\mu^\text{T} x$ while minimising risk, defined as the portfolio variance $x^\text{T}\Sigma x$, where $\mu$ is the vector of expected returns for each asset $i$ computed by Eq. \ref{eq:3}, $\Sigma$ is the covariance matrix calculate by Eq. \ref{eq:4} and Eq. \ref{eq:5}, and $x$ is the vector 
of asset allocations measured as fractions of the total budget. Therefore, the problem of finding the optimal portfolio is reduced to find the $x$ vector that maximises the following objective function: 
\begin{equation}
  \label{eq:6}
  \mathcal{L}(x) = \mu^\text{T} x - q x^\text{T}\Sigma x
\end{equation}
where $q$ is a parameter that determines the trade-off between risk and return.  In a realistic scenario, budget constraints and non-negative 
$x_i$ values (only buying) must be considered. Therefore, in the general case, the problem is formulated as follows:
\begin{equation}
\begin{aligned}
  \label{eq:7}
  \max_{x} \mathcal{L}(x): & \max_{x}\left(\mu^\text{T} x - q x^\text{T}\Sigma x \right) \\ 
  \text{s.t.} & \quad \sum_{i=1}^{N} x_i = 1 
\end{aligned}
\end{equation}
In section \ref{sec:literature1} we will review how this problem can be mapped into a QUBO problem and solved using quantum algorithms.

@@

The purpose of option pricing is to estimate Eq. \ref{eq:8} at the expiration date and then, determine the \textbf{fair value}, which indicates the amount of money an 
investor should pay to enter the option contract today, given the current market conditions. 
The challenging part is to compute the expected value of the option payoff, which is usually done using Monte Carlo methods. In section \ref{sec:literature2} we will discuss the quantum approach to 
this problem.

@@

The starting point is to define a cost function
\begin{equation}
  \label{eq:9}
  C(\boldsymbol{\theta}) =\langle\psi(\boldsymbol{\theta})|H|\psi(\boldsymbol{\theta})\rangle
\end{equation}
where $H$ is the Hamiltonian of our problem, that can be written as the weighted sum of Pauli operators \cite{Tilly2022}, such
\begin{figure*}[ht!]
  \begin{center} 
  \includegraphics[width=0.85\textwidth]{QAOA-Blekos.pdf}
  \caption{\label{fig:qaoa} Quantum Approximate Optimisation Algorithm (QAOA). Adapted from \cite{Blekos2024}.}
  \end{center}
\end{figure*}
that: 
\begin{equation}
  \label{eq:10}
  \mathcal{H} = \sum_{i} c_i P_i
\end{equation}
where $P_i \in \{I, X, Y, Z\}^{\otimes N}$ are called Pauli strings and $N$ is the number of qubits, and $c_i$ are the coefficients of the Pauli strings.  Additionally, $\boldsymbol{\theta}$ is a set of trainable parameters for some trial state $|\psi(\boldsymbol{\theta})\rangle = U(\boldsymbol{\theta}) |0\rangle ^{\otimes n}$. So, Eq. \ref{eq:9} becomes:
\begin{equation}
  \label{eq:11}
  C(\boldsymbol{\theta}) =\sum_i c_i \langle \boldsymbol{0}|U(\boldsymbol{\theta})^\dagger P_i U(\boldsymbol{\theta})|\boldsymbol{0}\rangle
\end{equation}
for some ansatz $U(\boldsymbol{\theta})$, that can be written as
\begin{equation}
  \label{eq:12}
  U(\boldsymbol{\theta}) = \prod_{k=1}^{p} e^{-i\theta_k \mathcal{H}_k}
\end{equation}
where $\mathcal{H}_k$ are the terms of the Hamiltonian $\mathcal{H}$  and $p$ determines the precision of the approximation \cite{Cerezo2021}.

@@

The variational principle ensures that at some point the VQE algorithm will converge to:
\begin{equation}
  \label{eq:13}
  E_0 \leq \frac{\langle \psi(\boldsymbol{\theta})|\mathcal{H}|\psi(\boldsymbol{\theta})\rangle}{\langle \psi(\boldsymbol{\theta})|\psi(\boldsymbol{\theta})\rangle}
\end{equation}
where $E_0$ is the ground state energy of the Hamiltonian $\mathcal{H}$ and $|\psi(\boldsymbol{\theta})\rangle$ is some trial state.

@@

Furthermore, they used their methodology to price a portfolio of options. They used these results to experimentally prove the speedup of the QAE algorithm (see Fig. \ref{fig:portfolio-option}).

@@

VQAs have been successfully implemented for optimisation problems. The results show that the performance of these algorithms is highly dependent on the ansatz structure and hardware topology. Nevertheless, the performance of the VQAs is formidable even when no error mitigation protocol is considered, making them a good candidate for exploiting the computational power that NISQi devices offer to date.

@@

\bibliography{/home/pacman/Documents/git-repositories/quantum-computing/lecture-notes/bibliography.bib}

>>>>

\begin{abstract}
Quantum computing has the promise to address computational problems that are considered intractable with our classical computational models. Nevertheless, due to the current limitations of the available quantum hardware, we are still years away from achieving a universal fault-tolerant quantum computer. In the interim, the development of quantum algorithms for specific problems and their implementation on near-term quantum devices is a promising approach. In this work, we review the most promising quantum algorithms for financial applications, focusing on Portfolio Optimisation and Option Pricing. We describe the ideas behind important quantum algorithms such as Variational Quantum Algorithms (VQAs) and Quantum Amplitude Estimation (QAE) and their theoretical performance. We provide an overview of the hardware  implementations and discuss the results 
obtained by some study cases, comparing them with their classical counterparts. Finally, we discuss the future prospects of these algorithms in the financial industry.
  \\
  \textbf{Keywords:} Quantum algorithms, Quantum approximate optimization algorithm, Quantum amplitude estimation, Portfolio optimisation, Quantum computing.
\end{abstract}

@@

 Finally, Sec. \ref{sec:discussion} concludes this work and discusses the future prospects of quantum algorithms in the financial industry.

@@

A set of investments $x_i$ (given as a fraction of the total budget or number of asset units) allocated for each $i$th considered asset, 
is then defined. Therefore, an optimal method for asset allocation aims to maximise the portfolio return 
$\mu^\text{T} x$ while minimising risk, defined as the portfolio variance $x^\text{T}\Sigma x$, where $\mu$ is the vector of expected returns for each asset $i$ computed by Eq. \ref{eq:3}, $\Sigma$ is the covariance matrix calculate by Eq. \ref{eq:4} and Eq. \ref{eq:5}, and $x$ is the vector 
of asset allocations measured as fractions of the total budget. Therefore, the problem of finding the optimal portfolio is reduced to find the $x$ vector that maximises the following objective function: 
\begin{equation}
  \label{eq:6}
  \mathcal{L}(x) = \mu^\text{T} x - q x^\text{T}\Sigma x
\end{equation}
where $q$ is a parameter that determines the trade-off between risk and return.  In a realistic scenario, budget constraints and non-negative 
$x_i$ values (only buying) must be considered. Therefore, in the general case, the problem is formulated as follows:
\begin{equation}
\begin{aligned}
  \label{eq:7}
  \max_{x} \mathcal{L}(x): & \max_{x}\left(\mu^\text{T} x - q x^\text{T}\Sigma x \right) \\ 
  \text{s.t.} & \quad \sum_{i=1}^{N} x_i = 1 
\end{aligned}
\end{equation}
Although analytical solutions exist for particular cases, the general problem is unsolvable efficiently with classical methods, and 
heuristic methods such as Monte Carlo simulations are used to find approximate solutions \cite{Shadabfar2020}.

@@

The purpose of option pricing is to estimate Eq. \ref{eq:8} at the expiration date and then, determine the \textbf{fair value}, which indicates the amount of money an 
investor should pay to enter the option contract today, given the current market conditions. 
The challenging part is to compute the expected value of the option payoff, which is usually done using Monte Carlo methods \cite{Sood2019}. In section \ref{sec:literature2} we will discuss the quantum approach to 
this problem.

@@

The starting point is to define a cost function
\begin{equation}
  \label{eq:9}
  C(\boldsymbol{\theta}) =\langle\psi(\boldsymbol{\theta})|H|\psi(\boldsymbol{\theta})\rangle
\end{equation}
where $H$ is the Hamiltonian of our problem, that can be written as the weighted sum of Pauli operators \cite{Tilly2022}, such
\begin{figure*}[ht!]
  \begin{center} 
  \includegraphics[width=0.85\textwidth]{QAOA-Blekos.pdf}
  \caption{\label{fig:qaoa} Quantum Approximate Optimisation Algorithm (QAOA). Adapted from \cite{Blekos2024}.}
  \end{center}
\end{figure*}
that: 
\begin{equation}
  \label{eq:10}
  \mathcal{H} = \sum_{i} c_i P_i
\end{equation}
where $P_i \in \{I, X, Y, Z\}^{\otimes N}$ are called Pauli strings and $N$ is the number of qubits, and $c_i$ are the coefficients of the Pauli strings.  Additionally, $\boldsymbol{\theta}$ is a set of trainable parameters for some trial state $|\psi(\boldsymbol{\theta})\rangle = U(\boldsymbol{\theta}) |0\rangle ^{\otimes n}$. Therefore, Eq. \ref{eq:9} becomes:
\begin{equation}
  \label{eq:11}
  C(\boldsymbol{\theta}) =\sum_i c_i \langle \boldsymbol{0}|U(\boldsymbol{\theta})^\dagger P_i U(\boldsymbol{\theta})|\boldsymbol{0}\rangle
\end{equation}
for some ansatz $U(\boldsymbol{\theta})$, that can be written as
\begin{equation}
  \label{eq:12}
  U(\boldsymbol{\theta}) = \prod_{k=1}^{p} e^{-i\theta_k \mathcal{H}_k}
\end{equation}
where $\mathcal{H}_k$ are the terms of the Hamiltonian $\mathcal{H}$  and $p$ determines the precision of the approximation \cite{Cerezo2021}.

@@

The variational principle ensures that after enough iterations the VQE algorithm will converge to:
\begin{equation}
  \label{eq:13}
  E_0 \leq \frac{\langle \psi(\boldsymbol{\theta})|\mathcal{H}|\psi(\boldsymbol{\theta})\rangle}{\langle \psi(\boldsymbol{\theta})|\psi(\boldsymbol{\theta})\rangle}
\end{equation}
where $E_0$ is the ground state energy of the Hamiltonian $\mathcal{H}$ and $|\psi(\boldsymbol{\theta})\rangle$ is some trial state.

@@

Furthermore, they used their methodology to price a portfolio of options, and using these results, they experimentally proved the speedup of the QAE algorithm (see Fig. \ref{fig:portfolio-option}).

@@

VQAs have been successfully implemented for optimisation problems. The results show that the performance of these algorithms is highly dependent on the ansatz structure and hardware topology. Nevertheless, the performance of the VQAs is formidable even when no error mitigation protocol is considered, making them a good candidate for exploiting the computational power that NISQ devices offer to date.

@@

\bibliography{./bibliography.bib}
